---
title: "Tables in Word Test"
author: "Derek Chiu"
date: "June 17, 2015"
output: word_document
---

```{r}
library(pander)
library(knitr)
set.seed(555)
n <- 50
num.col <- 10
mdat <- matrix(rnorm(n), ncol = num.col, byrow = TRUE,
               dimnames = list(paste0("Row", 1:(n / num.col)),
                               paste0("Col", 1:num.col)))
```

The first table uses `kable()`, which gives an output with awkward spacing as it tries to fit the table without separations

```{r}
kable(mdat, caption = "Table 1")
```

The second table uses `pander()`, and the formatting is OK, but the columns get split.

```{r}
pander(mdat, caption = "Table 2")
```

The third table uses `pander()` again but now we specify **not** to split the table, and also round the values.
Note also that this chunk needs the option `results='asis'`.

```{r, results='asis'}
pander(mdat, caption = "Table 3", split.table = Inf, round = 2)
```

There's a clever workaround from [Stack Overflow](http://stackoverflow.com/questions/30866999/how-to-emphasize-column-names-header-when-using-pandoc-in-r) showing how to **bold** the column headers (which isn't provided in `pander` because the header automatically gets highlighted by the thicker separating line in the table). The last table does this:

```{r, results='asis'}
colnames(mdat) <- pandoc.strong.return(colnames(mdat))
pander(mdat, caption = "Table 4", split.table = Inf, round = 2)
```
